### 2021-10-18 text/outline/outline.tex
update text in theoretical section

We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method. Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.

@@

\blue{\textbf{Bias}: We use the following theoretical model to study bias of the thresholding method: 
$$
\begin{cases}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
p_i \sim \textrm{Bern}(\pi) \\
\ep_i, \tau_i \sim N(0,1) \\
p_i \indep \tau_i \indep \ep_i
\end{cases}.
$$}
We can solve for the limit in probability of $\hat{\beta}_m$. We have that $$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
$$\begin{cases}
\omega = \pi\left( 1 - \Phi(c - \beta_1^g - \beta_0^g) \right),\\ \zeta = (1-\pi)(1 - \Phi\left( c - \beta^g_0 \right)), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
\end{cases}$$
Let $\gamma: \R^4 \to \R$ be defined by
$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the attenuation function. Observe that (i) $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and (ii) $ \hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$ Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$. We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.

@@

To investigate the basic question of ``What is a good threshold selection strategy?'', we studied the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$. For simplicity, we set the perturbation probability $\pi$ to $1/2$. Additionally, without loss of generality, we set the target of inference $\beta^m_1$ to $1$. Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e. $$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results relating the limiting bias $b$ of $\hat{\beta}^m_1$ to $c_\textrm{bayes}$ and $\beta^g_1$, deferring proofs to the appendix.

\begin{itemize}
\item[1.] The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e. 
$$ \frac{\partial b( \beta^g_0, \beta^g_1, c, 1/2)}{\partial c} \vline_{c = c_\textrm{bayes}} = 0.$$
\item[2.] For certain values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left); for other values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
\item[3.] The limit of the bias in $c$ is $1/2$, i.e. 
$$\lim_{c \to \infty} b(\beta^g_0, \beta^g_1, c, 1/2) = 1/2.$$ In other words, we always can set the threshold to a large number and achieve a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels). This fact establishes an upper bound on the bias of thresholded regression (under optimal threshold selection).
\item[4.] For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
\item[5.] The bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix a threshold and scan the panels from left to right). This is consistent with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
\item[6.] For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$. Therefore, the thresholding method suffers from attenuation bias, a common phenomenon in errors-in-variables models.
\end{itemize}

These results are subtle, but we can distill them as follows. First, selecting a threshold that minimizes the bias is deceptively challenging, as there is no rule of thumb that we can apply universally for this purpose (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number''). Furthermore, the optimal threshold depends on unknown parameters. Finally, even \textit{if} we have selected a good threshold, the thresholding method incurs a strict attenuation bias.

\begin{center}
\textbf{Bias of large threshold strategy vs.\ $\pi$ (Panel b)}
\end{center}

Intrigued by the finding that selecting a large threshold when $\pi = 1/2$ yields a bias of $1/2$, we investigated the relationship between bias and other values of $\pi$ under the large threshold selection strategy.

@@

>>>>

We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method. Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.

@@

We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:

\begin{equation}\label{theoretical_model}
\begin{cases}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
p_i \sim \textrm{Bern}(\pi) \\
\ep_i, \tau_i \sim N(0,1) \\
p_i \indep \tau_i \indep \ep_i.
\end{cases}
\end{equation}
For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_0$ for $\beta^m_0$ is given by $$\hat{\beta}^m_0 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
$$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
$$\begin{cases}
\omega = \pi\left( 1 - \Phi(c - \beta_1^g - \beta_0^g) \right),\\ \zeta = (1-\pi)(1 - \Phi\left( c - \beta^g_0 \right)), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
\end{cases}$$
Let $\gamma: \R^4 \to \R$ be defined by
$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the \textit{attenuation function}. Observe that \begin{itemize}
\item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
\item[ii.] $\hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
\end{itemize}
Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$. We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.

@@

To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$. For simplicity, we set the perturbation probability $\pi$ to $1/2$. Additionally, without loss of generality, we set the target of inference $\beta^m_1$ to $1$. Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e. $$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results relating the limiting bias $b$ of $\hat{\beta}^m_1$ to $c_\textrm{bayes}$ and $\beta^g_1$. We defer proofs to the appendix.

\begin{itemize}
\item[1.] The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the bias function $b$ (Figure \ref{thresholding_theoretical}a, all panels), i.e. 
$$ \frac{\partial b( \beta^g_0, \beta^g_1, c, 1/2)}{\partial c} \vline_{c = c_\textrm{bayes}} = 0.$$
\item[2.] For certain values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left); for other values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
\item[3.] The limit of the bias in $c$ is $1/2$, i.e. 
$$\lim_{c \to \infty} b(\beta^g_0, \beta^g_1, c, 1/2) = 1/2.$$ In other words, we always can set the threshold to a large number and achieve a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels). This fact establishes an upper bound on the bias of thresholded regression (under optimal threshold selection).
\item[4.] For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
\item[5.] The bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix threshold and scan panels from left to right). This is consistent with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
\item[6.] For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$ (Figure \ref{thresholding_theoretical}a, all panels). Therefore, the thresholding method suffers from attenuation bias.% a common phenomenon in errors-in-variables models.
\end{itemize}

These results are subtle, but we can summarize them as follows. First, selecting a threshold that minimizes the bias is deceptively challenging, as there is no rule of thumb that we can apply universally (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number''). Second, even if we \textit{have} selected a good threshold, the thresholding method incurs a strict attenuation bias.

\begin{center}
\textbf{Generalizing to $\pi \in [0,1/2]$ (Panel b)}
\end{center}

Next, we generalize the expression for bias to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$ In other words, when the perturbation probability is equal to $\pi$, setting the threshold to a large number yields a bias of $\pi$ (Figure \ref{thresholding_theoretical}b). We can build intuition for this result by considering an extreme example. When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed. Therefore, in selecting a large threshold, one correctly classifies nearly all \textit{unperturbed} cells as \textit{unperturbed}; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly \textit{unperturbed} cells, resulting in a small bias.

@@

Finally, we study the variance of the thresholding estimator. Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
$$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity: 
$$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.

To more precisely characterize the bias-variance tradeoff of the thresholding method, we consider a slightly simpler model: we set $\beta^m_0$ and $\beta^g_0$ to zero in (\ref{theoretical_model}) and study the no-intercept least squares estimator
$$ \hat{\beta}^m_1 =  \frac{ \sum_{i=1}^n \hat{p}_i m_i }{ \sum_{i=1}^n \hat{p}_i^2}.$$ We uncover an exact bias-variance tradeoff as a function of the selected threshold (Figure \ref{thresholding_theoretical}c). The bias-variance analysis reveals that the best strategy for minimizing mean-squared error is to select a threshold that induces moderate bias. (A downside of this approach, of course, is that constructing valid confidence intervals becomes more challenging.)
